---
title: "2026-04-21_hw04"
author: "Chris"
date: "2026-04-21"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    number_sections: true
    css: style.css
---


```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(caret)
library(rpart)
library(rpart.plot)
library(party)
library(partykit)
library(randomForest)
library(ranger)

set.seed(202604)
```


# Question One

```{r}
df_college <- read.csv("data/College.csv", row.names = 1) |>
  janitor::clean_names()
```


## Part I

### Fit the Model

```{r}
set.seed(202604)
data_split <- initial_split(df_college, prop = 0.8)

df_col_train <- training(data_split)
df_col_test <- testing(data_split)
```


```{r}
set.seed(202604)
tree1 <- rpart(
  outstate ~ ., data = df_col_train,
  control = rpart.control(cp = 0.005)
)

cpTable1 <- tree1$cptable
print(cpTable1)
```

### xerror

```{r}
set.seed(202604)
min_err1 <- which.min(cpTable1[, "xerror"])
tree2 <- prune(tree1, cp = cpTable1[min_err1, "CP"])
rpart.plot(tree2)
```


### 1se rule

```{r}
set.seed(202604)
min_row <- which.min(cpTable1[, "xerror"])
xerror_min <- cpTable1[min_row, "xerror"]
xstd_min   <- cpTable1[min_row, "xstd"]

threshold_1se <- xerror_min + xstd_min
candidate_rows <- which(cpTable1[, "xerror"] <= threshold_1se)
row_1se <- candidate_rows[1]

cp_1se <- cpTable1[row_1se, "CP"]

tree_1se <- prune(tree1, cp = cp_1se)

rpart.plot(tree_1se)
```

I randomly split the data into a training set (80%) and a test set (20%). Using the training data, I first fit a relatively large regression tree with an initial complexity parameter of 0.005. I then examined the cp table and considered two pruning strategies: selecting the cp that minimized the cross-validation error, and selecting the cp based on the 1-SE rule. These two approaches produced trees with different levels of complexity. Since the 1-SE tree was simpler and easier to interpret, I chose it as the final regression tree. The resulting plot is shown below.  


## Part II


### Fit the Model

```{r}
set.seed(202604)
ctrl <- trainControl(method = "cv", number = 10)
rf_grid <- expand.grid(mtry = 1:(ncol(df_col_train) - 1), 
                       splitrule = "variance", 
                       min.node.size = 1:6)

fit_rf <- train(outstate ~ ., 
                data = df_col_train, 
                method = "ranger", 
                tuneGrid = rf_grid,
                trControl = ctrl, 
                importance = "permutation")


```


### Variable Importance

```{r}
set.seed(202604)
fit_rf_final <- ranger(
  outstate ~ .,
  data = df_col_train,
  mtry = fit_rf$bestTune[[1]],
  splitrule = "variance",
  min.node.size = fit_rf$bestTune[[3]],
  importance = "permutation",
  scale.permutation.importance = TRUE
)

barplot(
  sort(ranger::importance(fit_rf_final), decreasing = FALSE),
  las = 2,
  horiz = TRUE,
  cex.names = 0.7,
  col = colorRampPalette(colors = c("cyan", "blue"))(length(ranger::importance(fit_rf_final)))
)
```


### Test Error

```{r}
rf_pred <- predict(fit_rf_final, data = df_col_test)$predictions
test_mse <- mean((rf_pred - df_col_test$outstate)^2)
test_rmse <- sqrt(test_mse)
test_mse
test_rmse
```

























